---
title: "Toronto Neighbourhood Crime Rate Overview"
author: 
  - Huakun Shen
thanks: "Code and data are available at: https://github.com/HuakunShen/Toronto-Crime-Rate-Analysis"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "The crime rates in different regions of Toronto ranges largely. The crime rate has an increasing trend from 2014 to 2019, and a sudden drop in 2020 probably due to covid 19."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

# Introduction

Crime Rate is one of the biggest concerns about peoples live and safety. 
OpenDataToronto[@OpenDataToronto] contains many public available datasets
containing data about Toronto. One of the datasets called 
"Neighbourhood Crime Rates" contains the number of different types of crimes 
happened in Toronto from 2014 to 2020. The crime rates (out of 100,000)
of each type of crime are also reported.

In this paper, I did some basic analysis on the topic of Toronto's crime rate
in recent years using [@citeR]. The crime rate in Toronto has a increasing 
trend starting 2014, and had a sudden drop in 2020. The sudden drop can probably 
be explained by Covid 19. The hypothesis will be explained in Section 
\@ref(discussion).

Some other R packages were also used. 
[@janitor] was used to clean up data frame column names.
[@tidyverse], [@tidyr], [@dplyr] were used for data processing.
[@ggpubr] was used to organize plots
[@knitr1], [@knitr2], [@knitr3] were used to generate tables.
[@teetor2011cookbook] was used for R syntax.

Section \@ref(data) explains where the dataset is taken from, what limitations it
has, and how I pre-processed the dataset. Section \@ref(results) shows the 
results (tables and plots) used for analysis. Section \@ref(discussion) contains
my understanding to the dataset, analysis, and explains some findings I derived
from the data. Furthermore, \@ref(weaknesses-and-next-steps) contains some 
thoughts about what analysis can be done in the future on this topic.



```{r setup, include=FALSE}
# install.packages("opendatatoronto")
# install.packages("knitr")
# install.packages("janitor")
# install.packages("lubridate")
# install.packages("opendatatoronto")
# install.packages("tidyverse")
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("ggpubr")
library(knitr)
library(janitor)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggpubr)
```


# Data

The dataset was downloaded from OpenDataToronto[@OpenDataToronto], 
"Neighbourhood Crime Rates" dataset.
This dataset includes the Crime Data by Neighbourhood, from 2014 to 2020, 
including 140 neighbourhoods.

No data is 100% accurate and precise. Assuming that there is no human factor 
affecting data collection, the data collected in this dataset may still
be inaccurate due to various reasons. For example, if a 
race or a group of people doesn't like to deal with police, the region where 
they are a majority might have a lower crime reporting rate, and the crime
rate if that region won't reflect the actual crime rate well. For example, a 
region full of gangster may have very high crime rate but lower-than-the-actual
reporting rate.

The types of crimes include Assault, Auto Theft, Break and Enter, Robbery, 
Theft Over, Homicide and Shooting & Firearm Discharges. Crime rate per 100,000 
is also calculated based on the population estimates provided by Environic 
Analytics.

A limitation of this dataset is its bad design. There is no "year" column, 
instead, the year is included in column names. For example, Crime type "Robbery"
has the following related columns

- Robbery_2014
- Robbery_2015
- Robbery_2016
- Robbery_2017
- Robbery_2018
- Robbery_2019
- Robbery_2020
- Robbery_Rate2014      
- RobberyRate_2015
- Robbery_Rate2016
- Robbery_Rate2017
- Robbery_Rate2018
- Robbery_Rate2019
- Robbery_Rate2020

There are 104 columns in total while there are only 7 crime types. It's very
hard to do analysis on years with this dataset structure, for example, I can't 
do a crime rate by year on this.


```{r get package, echo = FALSE}
package_id <- "fc4d95a6-591f-411f-af17-327e6c5d03c7"
package <- show_package(package_id)

# get all resources for this package
resources <- list_package_resources(package_id)
 
# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
 
# load the first datastore resource as a sample
data <- filter(datastore_resources, row_number()==1) %>% get_resource()
```

Table \@ref(tab:original-dataset-preview) is a subset preview of the original 
dataset. It's impossible to show the entire table as there are too many columns.

```{r original-dataset-preview, echo=FALSE}
cols_to_show <- c("Neighbourhood", "Assault_2014", "Assault_2015")
kable(head(data[cols_to_show]), "pipe", caption = "Preview of a subset of the original dataset")
```

```{r echo = FALSE}
years <- c("2014", "2015", "2016", "2017", "2018", "2019", "2020")
categories <- c("Assault_","Assault_Rate","AutoTheft_","AutoTheft_Rate",
               "BreakAndEnter_","BreakAndEnter_Rate","Robbery_","Robbery_Rate",
               "TheftOver_","TheftOver_Rate","Homicide_","Homicide_Rate",
               "Shootings_","Shootings_Rate")
unique_neighbourhood <- unique(data$Neighbourhood)
unique_hood_id <- unique(data$Hood_ID)
```


```{r echo = FALSE}
categories <- c("Assault_","Assault_Rate","AutoTheft_","AutoTheft_Rate",
               "BreakAndEnter_","BreakAndEnter_Rate","Robbery_","Robbery_Rate",
               "TheftOver_","TheftOver_Rate","Homicide_","Homicide_Rate",
               "Shootings_","Shootings_Rate")
column.names <- c("hood_id", "year", "Neighbourhood", "F2020_Population_Projection", categories)
column.length <- length(column.names)
row.length <- length(years) * length(unique_hood_id)
mat <- matrix(ncol = column.length, nrow = 0)
df <- data.frame(matrix(ncol = column.length, nrow = 0))
colnames(df) <- column.names


for (hood_id in unique_hood_id) {
  hood_data <- data[data$Hood_ID == hood_id,]
  for (year in years) {
    v = c(hood_id, year, hood_data$Neighbourhood, hood_data$F2020_Population_Projection)
    for (category in categories) {
      col_name <- paste(category, year, sep="")
      val <- hood_data[[col_name]]
      if (is.null(val)) {
        v <- c(v, 0)
      } else {
        v <- c(v, val)
      }
    }
    df[nrow(df) + 1,] = v
  }
}
```


```{r Rename Columns, echo = FALSE}
for (category in categories) {
  df[[category]] = as.numeric(df[[category]])
}
df$F2020_Population_Projection = as.numeric(df$F2020_Population_Projection)
names(df)[names(df) == 'Assault_'] <- 'Assault'
names(df)[names(df) == 'AutoTheft_'] <- 'AutoTheft'
names(df)[names(df) == 'BreakAndEnter_'] <- 'BreakAndEnter'
names(df)[names(df) == 'Robbery_'] <- 'Robbery'
names(df)[names(df) == 'TheftOver_'] <- 'TheftOver'
names(df)[names(df) == 'Homicide_'] <- 'Homicide'
names(df)[names(df) == 'Shootings_'] <- 'Shootings'
```

To solve this problem, I pre-processed the data to make it have a "year" column.
The resulting dataset has the structure in table \@ref(tab:cleaned-data-preview).


```{r cleaned-data-preview, echo=FALSE}
cols_to_show <- c("Neighbourhood", "year", "Shootings_Rate", "Robbery_Rate", "Homicide_Rate")
knitr::kable(head(df[cols_to_show]), "pipe", caption="Cleaned Table (Subset)")
```


# Results

```{r calculate average crime rate by neighbourhood, echo = FALSE}
df_gb <- df %>% group_by(Neighbourhood)
neighbourhood_mean_crime_rate <- df_gb %>% summarise(
  Assault_Rate=mean(Assault_Rate),
  AutoTheft_Rate=mean(AutoTheft_Rate),
  BreakAndEnter_Rate=mean(BreakAndEnter_Rate),
  Robbery_Rate=mean(Robbery_Rate),
  TheftOver_Rate=mean(TheftOver_Rate),
  Homicide_Rate=mean(Homicide_Rate),
  Shootings_Rate=mean(Shootings_Rate)
)
```

## Mean Crime Rate

### Mean Crime Rate By Neighbourhood

The mean crime rate is calculated for every neighbourhood by grouping-by the
"Neighbourhood" column.

Here is a preview of the resulting dataset.

```{r echo=FALSE}
cols_to_show <- c("Neighbourhood", "Shootings_Rate", "Robbery_Rate", "Homicide_Rate")
kable(head(neighbourhood_mean_crime_rate[cols_to_show]), "simple", caption = "Average Crime Rate By Neighbourhood (Subset)")
```


```{r shootings-rate-by-neighbourhood, fig.cap="Shootings Rate By Neighbourhood", fig.width=14,, fig.height=10, echo = FALSE}
ggplot(neighbourhood_mean_crime_rate,
       aes(x = reorder(Neighbourhood, -Shootings_Rate), y = Shootings_Rate)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("Neightbourhood") + ylab("Shootings Rate") +
  ggtitle("Neighbourhood Shootings Rate (out of 100,000)")
# ggsave("output_images/neightbourhood-Shootings-Rate.png")
```

```{r calculate average crime rate by year, echo = FALSE}
df_gb <- df %>% group_by(year)
mean_crime_rate_by_year <- df_gb %>% summarise(
  Assault_Rate=mean(Assault_Rate),
  AutoTheft_Rate=mean(AutoTheft_Rate),
  BreakAndEnter_Rate=mean(BreakAndEnter_Rate),
  Robbery_Rate=mean(Robbery_Rate),
  TheftOver_Rate=mean(TheftOver_Rate),
  Homicide_Rate=mean(Homicide_Rate),
  Shootings_Rate=mean(Shootings_Rate)
)
mean_crime_rate_by_year <- mean_crime_rate_by_year[order(mean_crime_rate_by_year$year),]
```

### Mean Crime Rate By Year

The mean crime rate is calculated for every year by grouping-by the"year" column.

Here is a preview of the resulting dataset.

```{r echo=FALSE}
cols_to_show <- c("year", "Shootings_Rate", "Robbery_Rate", "Homicide_Rate")
knitr::kable(head(mean_crime_rate_by_year[cols_to_show]), "pipe", caption="Average crime Rate By Year (Subset)")
```

```{r Plot Yearly crime Rate, echo = FALSE}
assault_rate_plt <- ggplot(mean_crime_rate_by_year,
       aes(x = year, y = Assault_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("Assault Rate") +
  ggtitle("Assault Rate by year (out of 100,000)")
# ggsave("output_images/assault-rate-by-year.png")
auththeft_rate_plt <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = AutoTheft_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("AutoTheft Rate") +
  ggtitle("AutoTheft Rate by year (out of 100,000)")
# ggsave("output_images/autotheft-rate-by-year.png")
breakAndEnter_plt <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = BreakAndEnter_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("BreakAndEnter Rate") +
  ggtitle("BreakAndEnter Rate by year (out of 100,000)")
# ggsave("output_images/breakandenter-rate-by-year.png")
robbery_rate_plot <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = Robbery_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("Robbery Rate") +
  ggtitle("Robbery Rate by year (out of 100,000)")
# ggsave("output_images/robbery-rate-by-year.png")
theftOver_rate_plt <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = TheftOver_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("TheftOver Rate") +
  ggtitle("TheftOver Rate by year (out of 100,000)")
# ggsave("output_images/theftover-rate-by-year.png")
homicide_rate_plt <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = Homicide_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("Homicide Rate") +
  ggtitle("Homicide Rate by year (out of 100,000)")
# ggsave("output_images/homicide-rate-by-year.png")
shootings_rate_plt <- ggplot(mean_crime_rate_by_year, 
       aes(x = year, y = Shootings_Rate)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Year") + ylab("Shootings Rate") +
  ggtitle("Shootings Rate by year (out of 100,000)")
# ggsave("output_images/shootings-rate-by-year.png")
```

Figure \@ref(fig:crime-rate-by-year) shows the crime rates of different
types from 2014 to 2020. 

```{r crime-rate-by-year, fig.cap="Crime Rate By Year", fig.width=16, fig.height=16, echo=FALSE}
ggarrange(assault_rate_plt,
          auththeft_rate_plt,
          breakAndEnter_plt,
          robbery_rate_plot,
          theftOver_rate_plt,
          homicide_rate_plt,
          shootings_rate_plt,
          labels = c("A", "B", "C", "D", "E", "F", "G"),
          ncol = 2, nrow = 4)
```


# Discussion

## Crime Rate in different Neighbourhoods

Figure \@ref(fig:shootings-rate-by-neighbourhood) shows the Shootings Rate for 
different neighbourhoods. The crime rates range greatly across different regions.
The region with the highest crime rate has approximately 65 shootings per 100,000
people. The region with the least shootings has close to 0 shooting.

If a person were pick a residence, it's better for them to pick a region with 
lower crime rate, especially shootings rate.

## Mean Crime Rate By Year

Figure \@ref(fig:crime-rate-by-year) shows the crime rate of different types by
year.
We can observe that, all crime types except for robbery has an increasing rate. 
Meaning that The overall crime rate has been increasing in Toronto in recent years.

An interesting observation is that all crime types except for Robbery and 
Auto Theft have a increasing trend if crime rate but a sudden drop in 2020. 
Given that the data is correct, the sudden drop is a surprise.

Covid 19 is undoubtedly one of the most unexpected variable in year 2020, and it
can be used to explain the sudden drop of crime rates in 2020.

The sudden drop of crime rate of Assault, Break-and-Enter, Theft Over, Homicide, 
Shootings can be explained by the stay-at-home order caused by the pandemic.

Although many people ignored the order, the population going out is still 
reduced by work-from-home, prohibition to dine-in. As people tend to stay home 
for longer time, crimes like assualt, theft over, break-and-enter, homicide and 
shootings which usually happen outside of home tend to occur less frequently.

The exception is Auto Theft crime, the only crime rate increased in 2020. 
It can also be explained by the same reason. As people stay home for longer,
their vehicles are also left alone for longer, giving the thefts more chances to
steal a vehicle.

Robbery is the only crime that has a constant dropping rate. 2020 has a 
exceptionally large drop compared to previous years, which could be explained 
by the same reason.



## Weaknesses and next steps

### Neighbourhood Safety

I only generated the shootings rate histogram for neighbourhoods (which is the 
most critical type of crime). To better understand the safety in different 
neighbourhoods, I will need to generate a histogram for each crime type.
As there are 140 neighbourhoods, plotting a histogram isn't intuitive enough to
visualize the crime rate in different regions. The next step could be plotting a
heat map, which is a more intuitive visualization. 

It's also interesting that the crime rate ranges largely in different regions.
The next step could be finding out what are the causes of the high crime rate in
some regions. For example, housing price, rental price, race of residents, 
education level of residents, population density, income level of residents, etc.
The dataset used in this paper is not enough, more data will be required to do
such analysis.

### Cause of Crime Rate Drop In 2020

Based on data in the dataset used in this paper, it's not rigorous enough to 
conclude that Covid 19 reduces crime rate. The crime rates were reported annually,
it should be more accurate if we can use monthly reported crime data and covid 
counts. 




\newpage

# References














